flowCore: data structures package for flow cytometry data

### 2007-10-01 inst/doc/HowTo-flowCore.Rnw
updated vignette to include filterSet examples. removed extraneous print statement and fixed two method signatures for filterSet operations.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@27678 bc3139a8-67e5-0310-9ffc-ced21a209358

>>>>

\section{filterSet: Organizing Filtering Strategies}

\subsection{Building Filter Sets}

In addition to allowing for sequential filtering, \Rpackage{flowCore} also provides a \Rclass{filterSet} object that serves as an analogue to \Rclass{flowSets} for filters. The primary use of this object is to organize and manipulate complex gating strategies. For example, recall from the filtering section we had two gates, a ``morphologyGate'' and a ``Fluorescence Region'' gate that we used to demonstrate the various logical operators available for gates. To recreate this example as a \Rclass{filterSet}, we would do the following:

<<FilterSet1, echo=true, results=verbatim>>=
fset1 = filterSet(
	rectangleGate(filterId="Fluorescence Region","FL1-H"=c(50,100),"FL2-H"=c(50,100)),
	norm2Filter(filterId="Morphology Gate","FSC-H","SSC-H",scale=2),
	~ `Fluorescence Region` & `Morphology Gate`,
	~ `Fluorescence Region` | `Morphology Gate`,
	Debris ~ ! `Morphology Gate`,
	~ `Fluorescence Region` %&% `Morphology Gate`
)
fset1

There are two features of note in \Rfunction{filterSet}, which can also take a single list argument for programmatic creation of \Rclass{filterSet} objects. The first is that there is a formula interface for the creation of \Rclass{filter} objects from operators. The formula interface can be used with or without a left-hand side, which specifies an optional filter identifier. If the filter identifier is not specified, as is the case with all but the ``Debris'' gate in the example above the usual default filter identifier is constructed. Non-formula gates can also have an optional name specified which overrides the filter identifier specified in the gate constructor. This is discouraged at this time as it leads to confusion when creating new filters.

\subsection{Manipulating Filter Sets}

Manipulating a \Rclass{filterSet} is done using the normal list element replacement methods, though we do provide a special case where replacing the ``""'' element or the ``NULL'' element causes the \Rclass{filterSet} to use the ``filterId'' slot of the incoming filter to choose a name. Similarly, specifying an identifier will override any ``filterId'' slot in the original filter.

Additionally, there are several convenience functions and methods available for manipulating a \Rclass{filterSet}. The \Rfunction{names} function provides a list of filter identifiers in an arbitrary order. The \Rfunction{sort} lists the filter identifiers according to a topological sort of the filter dependencies such that parent filters are always evaluated before their children. Filters without dependencies are provided in an arbitrary relative ordering. To obtain the adjacency matrix for this sorting, the option ``dependencies=TRUE'' supplied to the \Rfunction{sort} function will attach an ``AdjM'' attribute to the resulting character vector.

\subsection{Using Filter Sets}

Though they are not explicitly filters, meaning that they cannot play a role in the construction of subfilters via operators, \Rclass{filterSet} objects do define \Rfunction{filter} and \Rfunction{split} methods that allow for the two most common use cases. Additionally, a filter can be selected from the \Rclass{filterSet} using ``[['' to perform a \Rfunction{Subset} operation if desired.

For standard filtering operations, the \Rclass{filterSet} can yield better performance than a manual filtering operation because it ensures that each parent filter is only evaluated once per call to \Rfunction{filter} so that if many subfilters rely on a particular, likely expensive, filtering operation they use the original result rather than recalculating each time. The individual \Rclass{filterResult} objects may be later extracted for further analysis of desire, making the \Rclass{filterSet} the logical choice for most filtering operations. Additionally, though not presently implemented, it can allow visualization methods to make more intelligent choices about display when rendering gating strategies because the disposition of the entire filtering graph is known and the filter results are already available.

Like any other filtering operation, \Rfunction{summary} methods are also available to collect the usual sorts of count statistics for a filtering operation:

<<FilterSet2, echo=TRUE, verbatim=TRUE>>=

f = filter(fs,fset1)
f
@

When splitting a \Rclass{flowFrame} using \Rfunction{split}, we have two options. The first is to include all of the resultant subsets in the usual way:

<<FilterSet3, echo=TRUE, verbatim=TRUE>>=

split(fs[[1]],fset1,flowSet=TRUE)

While this provides access to all subsets, for automated analysis we will often only wish to work with the ``leaf'' gates (i.e. those without children of their own). In this case, we can specify ``drop=TRUE'' to obtain only the leaf gates:

<<FilterSet4, echo=TRUE, verbatim=TRUE>>=

split(fs[[1]],fset1,drop=TRUE,flowSet=TRUE)

Note that in both cases, we use the ``flowSet=TRUE'' option that has been added to the \Rfunction{split} function to automatically create a \Rclass{flowSet} object from the resulting \Rclass{flowFrame} objects. For \Rfunction{filterSet} objects, the \Rfunction{split} function can take advantage of the added structure in this case to provide some information about the filtering operation itself in the \Rclass{flowSet}'s metadata as well.
